Properties of Winning Iterated Prisoner's Dilemma Strategies.

### 2024-08-07 paper/main.tex
update paper

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades, from the celebrated performance of Tit for Tat to the
introduction of the zero-determinant strategies and the use of sophisticated learning
structures such as neural networks. Many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies in a small number of tournaments, casting doubt on the generalizability
of conclusions. In this work, we analyze a large collection of \numberofstrategies
strategies in thousands of computer tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features. 
We find that although a strategy's high ranking in a specific environment
does not guarantee success across different tournament types, certain strategies
consistently perform well across various settings. Notably,
strategies designed or trained for noise are the strategies that generalise
better—an observation well established in AI/ML literature where adding noise to
training data enhances model robustness.
We identify several crucial properties
heavily influencing the success of strategies, refining Axelrod's described
properties in light of encounters with more diverse opponents to: being
cooperative, responsive to provocation, generous yet slightly envious, clever,
and adaptable to the prevailing environment. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities.
The features of high performing strategies help cast some light on why
strategies such as Tit For Tat performed historically well in tournaments and
why zero-determinant strategies typically do not fare well in tournament
settings. Furthermore, our findings have implications for the future training of
autonomous agents, as understanding the crucial features for incorporation into
these agents becomes essential.
\end{abstract}

\section{Background}

The Iterated Prisoner's Dilemma (\IPD) is a repeated two-player game that models
behavioral interactions, specifically interactions where self-interest clashes
with collective interest. In each turn of the game, both players simultaneously
and independently decide between cooperation (\(C\)) and defection (\(D\)). This
decision is made with the memory of all prior interactions. The payoffs for each
player at each turn are influenced by their own choice and the choice of the
other player. To this end, the payoffs of the game are defined by

@@

Conceptualizing strategies and understanding the best way to play the game have
been of interest to the scientific community since the formulation of the game
in 1950~\cite{Flood1958}. Following Axelrod's computer tournaments in the
1980s~\cite{Axelrod1980a, Axelrod1980b}, round-robin computer tournaments became
a common evaluation technique for newly designed strategies.
The winner of both of Axelrod's tournaments~\cite{Axelrod1980a, Axelrod1980b}
was the simple strategy Tit For Tat (\TFT{}). \TFT{} cooperates on the first turn
and thereafter copies the previous action of its opponent, retaliating against
defections with a defection and forgiving a defection if followed by
cooperation. Axelrod concluded that the strategy's robustness was due to four
properties, which he adapted into four suggestions for success in an \IPD{}
tournament:

@@

Finally, others introduced strategies deviating completely from the originally
suggested properties of success. For example, a set of ``envious'' Iterated
Prisoner's Dilemma (\IPD) strategies were introduced, called zero-determinant
strategies (ZDs), in~\cite{Press2012}. These strategies attempt to force a
linear relationship between stationary payoffs against other memory-one
opponents, potentially ensuring that they receive a higher average payout. While
ZDs were introduced with a small tournament in which some were reportedly
successful \cite{Stewart2012}, this result has not generally held in future
work~\cite{mathieu2017}. Furthermore, in~\cite{Harper2017}, a series of ``clever''
strategies trained using reinforcement learning were introduced. These
strategies were trained using lookup tables~\cite{Axelrod1987}, hidden Markov
models~\cite{Harper2017}, and finite-state automata~\cite{Miller1996}, on
a set of 170 strategies.

@@

It is important to note that we do not assert the existence of a single
best-performing strategy across all tournaments or tournament types. On the
contrary, our work demonstrates that such a strategy does not always exist
(notwithstanding a few strategies with broadly high performance). The primary
objective of this paper, presented in the latter parts of the paper, is to
continue the discussion on the properties of successful strategies, a
conversation started by Axelrod. 
The results of our analysis conclude that the properties of a successful
strategy in the Iterated Prisoner's Dilemma (\IPD{}) are:

@@

We believe that the discussion on the properties of winning strategies holds
significant importance. It aims to provide guidance to researchers designing new
strategies and those training strategies. Specifically, much like the recognized
value of diversity in training datasets, such as variations in image perspective,
skin color, etc., are critical in training accurate and generalizable machine
learning models, we show that diversity in the population of opponent strategies
is of paramount importance in the construction and evaluation of game theory
strategies. As AI agents start to interact as scale, they will benefit from
exposure to a wide variety of alternative agents. Moreover, conducting a similar
analysis can shed light on already trained strategies, aiding in understanding
the key features they have autonomously developed during their training
processes.

@@

\section{Data collection}\label{section:data_collection}

@@

SSE & A measure of how far a strategy is from ZD behaviour & method described in~\cite{Knight2019} & float & 0 & 1 \\

@@

    turns. The SSE (introduced in~\cite{Knight2019}) shows how close a strategy

@@

This might also be the reason why ZDs fail to appear in the top ranks—they
attempt to exploit all opponents and cannot actively adapt back to mutual
cooperation against stronger strategies, which requires a deeper memory. It's
worth noting that ZDs tend to perform poorly in population games for a similar
reason: they aim to exploit other players using ZDs, failing to form a
cooperative subpopulation~\cite{Knight2017evolution}. This makes them effective
invaders but poor at resisting invasion.

@@

This manuscript explores the performance of \numberofstrategies strategies in
the \IPD{} in thousands of computer tournaments. The collection of computer
tournaments presented here is the largest and most diverse in the literature.
The \numberofstrategies strategies are drawn from \axelrod{} library and include
strategies from the \IPD{} literature. The computer tournaments encompass four
different types. So, what is the best way to play the \IPD{}? And is there a
single dominant strategy for the \IPD{}?
A strategy ranking highly in a
specific environment did not guarantee its success over different tournament
types, with a few exceptions - strategies that generalize better. Already
well-known in the AI/ML literature, adding noise to training data leads to more
robust models. We see that clearly here, where the strategies trained for noise
(or designed for noise) tend to be better generalists. There were instances
where a few strategies trained in narrow conditions outperformed more generalist
strategies, as they tend to overfit. However, the strategies trained with noise
perform well in general, whilst the strategies trained specifically on no noise
or small subpopulations do not.

>>>>

\begin{abstract}
Researchers have explored the performance of Iterated Prisoner's Dilemma strategies
for decades, from the celebrated performance of Tit for Tat to the
introduction of the zero-determinant strategies and the use of sophisticated learning
structures such as neural networks. Many new strategies have been introduced and tested
in a variety of tournaments and population dynamics. Typical results in the literature,
however, rely on performance against a small number of somewhat arbitrarily selected
strategies, casting doubt on the generalizability
of conclusions. In this work, we analyze a large collection of \numberofstrategies
strategies in thousands of computer tournaments, present the top performing strategies across multiple
tournament types, and distill their salient features.
The results show that there is not yet a single
strategy that performs well in diverse Iterated Prisoner's Dilemma scenarios,
nevertheless there are several properties that heavily influence the best performing
strategies. This refines the properties described by Axelrod in light of
recent and more diverse opponent populations to: be nice, be provocable and generous,
be a little envious, be clever, and adapt to the environment. More precisely,
we find that strategies perform best when their probability of cooperation
matches the total tournament population's aggregate cooperation probabilities.
The features of high performing strategies help cast some light on why
strategies such as Tit For Tat performed historically well in tournaments and
why zero-determinant strategies typically do not fare well in tournament
settings.
\end{abstract}

\section{Introduction}

The Iterated Prisoner's Dilemma (\IPD) is a repeated two-player game that models
behavioral interactions, specifically interactions where self-interest clashes
with collective interest. It encompasses a wide range of social and
biological phenomena. In each turn of the game, both players simultaneously
and independently decide between cooperation (\(C\)) and defection (\(D\)). This
decision is made with the memory of all prior interactions. The payoffs for each
player at each turn are influenced by their own choice and the choice of the
other player. To this end, the payoffs of the game are defined by

@@

Conceptualizing strategies and understanding the best way to play the game have
been of interest to the scientific community since the formulation of the game
\cite{Axelrod1980a, Beaufils1997, Press2012, Tzafestas2000, Hilbe2017,
Glynatsi2020, Murase2021, Schmid2022, Chen2023}.
This extends to both tournament settings and population dynamics.
Computer tournaments became a common evaluation technique for newly designed
strategies following Axelrod's computer tournaments in the
1980s~\cite{Axelrod1980a, Axelrod1980b}.
The winner of both of Axelrod's tournaments~\cite{Axelrod1980a, Axelrod1980b}
was the simple strategy Tit For Tat (\TFT{}). \TFT{} cooperates on the first turn
and thereafter copies the previous action of its opponent, retaliating against
defections with a defection and forgiving a defection if followed by
cooperation. Axelrod concluded that the strategy's robustness was due to four
properties, which he adapted into four suggestions for success in an \IPD{}
tournament:

@@

Finally, others introduced strategies deviating completely from the originally
suggested properties of success. For example, a set of ``envious'' Iterated
Prisoner's Dilemma (\IPD) strategies were introduced, called zero-determinant
strategies (ZDs), in~\cite{Press2012}. These strategies attempt to force a
linear relationship between stationary payoffs against their
opponents, potentially ensuring that they receive a higher average payout. While
ZDs were introduced with a small tournament in which some were reportedly
successful \cite{Stewart2012}, this result has not generally held in future
work~\cite{mathieu2017}. Furthermore, in~\cite{Harper2017}, a series of ``clever''
strategies trained using reinforcement learning were introduced. These
strategies were trained using lookup tables~\cite{Axelrod1987}, hidden Markov
models~\cite{Harper2017}, and finite-state automata~\cite{Miller1996}, on
a set of 170 strategies.

@@

It is important to note that we do not assert the existence of a single
best-performing strategy across all tournaments or tournament types. On the
contrary, our work demonstrates that such a strategy does not exist
(notwithstanding a few strategies with broadly high performance). The primary
objective of this paper, presented in the latter parts of the paper, is to
continue the discussion on the properties of successful strategies, a
conversation started by Axelrod. 
The results of our analysis conclude that the properties of a successful
strategy in the Iterated Prisoner's Dilemma (\IPD{}) are:

@@

We believe that the discussion on the properties of winning strategies holds
significant importance. It aims to provide guidance to researchers designing new
strategies and those training strategies. Specifically, much like the recognized
value of diversity in training datasets, such as variations in image perspective,
skin color, etc., are critical in training accurate and generalizable machine
learning models, we show that diversity in the population of opponent strategies
is of paramount importance in the construction and evaluation of game theory
strategies. Moreover, conducting a similar
analysis can shed light on already trained strategies, aiding in understanding
the key features they have autonomously developed during their training
processes.

@@

\section{Model}\label{section:data_collection}

@@

SSE & A measure of how far a strategy is from ZD behaviour & method described in~\cite{Knight2024} & float & 0 & 1 \\

@@

    turns. The SSE (introduced in~\cite{Knight2024}) shows how close a strategy

@@

This might also be the reason why ZDs fail to appear in the top ranks—they
attempt to exploit all opponents and cannot actively adapt back to mutual
cooperation against stronger strategies, which requires a deeper memory. It's
worth noting that ZDs tend to perform poorly in population games for a similar
reason: they aim to exploit other players using ZDs, failing to form a
cooperative subpopulation~\cite{Knight2018evolution}. This makes them effective
invaders but poor at resisting invasion.

@@

This manuscript explores the performance of \numberofstrategies strategies in
the \IPD{} in thousands of computer tournaments. The collection of computer
tournaments presented here is the largest and most diverse in the literature.
The \numberofstrategies strategies are drawn from \axelrod{} library and include
strategies from the \IPD{} literature. The computer tournaments encompass four
different types. So, what is the best way to play the \IPD{}? And is there a
single dominant strategy for the \IPD{}? There was not a single strategy within
the collection of \numberofstrategies strategies that managed to perform well in
all the tournament variations it competed in. A strategy ranking highly in a
specific environment did not guarantee its success over different tournament
types, with a few exceptions - strategies that generalize better. Already
well-known in the AI/ML literature, adding noise to training data leads to more
robust models. We see that clearly here, where the strategies trained for noise
(or designed for noise) tend to be better generalists. There were instances
where a few strategies trained in narrow conditions outperformed more generalist
strategies, as they tend to overfit. However, the strategies trained with noise
perform well in general, whilst the strategies trained specifically on no noise
or small subpopulations do not.
